---
title: "Assigning ASVs with DADA2"
author: "Kailyn Hanke"
date: "2025-03-11"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center", 
                      # Send figures generated in this file to this folder below
                      fig.path = "../figures/02_AssignASVs/")
```

In this analysis document I will be assigning ASVs with DADA2. I will be using data from a 2020 paper titled: "Tracking the Dairy Microbiota from Farm Bulk Tank to Skimmed Milk Powder" (@mchughTrackingDairyMicrobiota2020). I am assigning ASVs to eventaully answer my broad research question: How does microbial diversity change throughout the skim milk powder manufacturing process?

# Goals 

1. Infer an error model for in the filtered sequences, separately on forward and reverse reads.
2. Assign ASVs on both forward and reverse reads separately by applying the error model.
3. Merge forward and reverse ASVs into "contiguous ASVs".
4. Generate the first draft of ASV count table.
5. Quality Trimming of ASV lengths.
6. Remove chimeras. 
7. Assign Taxonomy with Silva Database. 
8. Write out relevant files: `asv_table`, `asvs_fasta`, `tax_table`, and `sample_data`.

## Input 

1. Filtered fastq files generated from `01_QualityTrimming.Rmd`.
2. Sample Name vector.

## Output 

1. ASV Count Table: `asv_table` (with and without sequence names)
2. ASV fasta file: `asvs_fasta` for building a phylogenetic tree at a later step.
3. Taxonomy Table  `tax_table`
4. Sample Information: `sample_data`  track the reads lots throughout DADA2 workflow. 

# Set up the Environment 

## Set Seed & Threads

```{r set-seed}
# Set the seed for reproducibility
set.seed(111111)

# Let's make a parameter to set the number of threads 
n_threads = 35
```

## Timing of this script

Let's record how long this file took to run on the class server, which we will record at the end of the script. 

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Load Packages 
```{r load-packages}
# Efficient package loading with pacman 
pacman::p_load(tidyverse, devtools, dada2, 
               patchwork, DT, dplyr, install = FALSE)
```
## Load Filtered Fastq Files 

```{r load-filtered-fastqs}
# Place filtered seq files into a variable 
filtered_fastqs_path <- "data/01_DADA2/02_filtered_fastqs"

# Intuition check:
filtered_fastqs_path

# Create Forward vector 
filtered_forward_reads <- 
  list.files(filtered_fastqs_path, pattern = "_1_filtered.fastq.gz",
             full.names = TRUE) 
# Check 
filtered_forward_reads[1:5]

# Reverse vector 
filtered_reverse_reads <- 
    list.files(filtered_fastqs_path, pattern = "_2_filtered.fastq.gz",
             full.names = TRUE)

# Check 
filtered_reverse_reads[1:5]
```

## Sample Names 
```{r sample-names}
# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(filtered_forward_reads), "_"), `[`,1) 

# Intuition Check 
head(sample_names)
```

## Learn the Errors


```{r learn-errors-MiSeq, fig.width=12, fig.height=8}
# Forward Reads 
error_forward_reads <- 
  learnErrors(filtered_forward_reads, multithread = n_threads)

# Forward Error Plot 
forward_error_plot <- 
  plotErrors(error_forward_reads, nominalQ = TRUE) + 
  labs(title = "Forward Reads: Error Model")

# Reverse Reads 
error_reverse_reads <- 
  learnErrors(filtered_reverse_reads, multithread = n_threads)

# Reverse Error Plot 
reverse_error_plot <- 
  plotErrors(error_reverse_reads, nominalQ = TRUE) + 
  labs(title = "Reverse Reads: Error Model")

# Look at the plots together 
forward_error_plot + reverse_error_plot
```
# Infer ASVs


```{r infer-ASVs}
# Infer ASVs on the forward sequences
dada_forward <- 
  dada(filtered_forward_reads, 
       err = error_forward_reads,
       multithread = n_threads) 

# Take a look at the data
# What type of data structure is it? 
typeof(dada_forward) # It's a list 
length(dada_forward) # How big is it? One per sample!

# What doees it look like for each sample?  
dada_forward$`ERR3141386_1_filtered.fastq.gz`

# Reverse ASVs
dada_reverse <- 
  dada(filtered_reverse_reads,
       err = error_reverse_reads ,
       multithread = n_threads)

# Check data 
dada_reverse[30]
```

# Merge Forward and Reverse ASVs 


```{r merge-ASVs}
merged_ASVs <- 
  mergePairs(dada_forward, filtered_forward_reads,
             dada_reverse, filtered_reverse_reads,
             verbose = TRUE)

# Evaluate the data output 
typeof(merged_ASVs) # A list
length(merged_ASVs) # Length of the number of samples!
head(names(merged_ASVs)) # Here, we can access our current sample names

# Inspect further for each sample
#head(merged_ASVs, n = 2) # A dataframe for each sample
# We have a dataframe in each part of our list! What are in the columns? 
glimpse(merged_ASVs$`ERR3141386_1_filtered.fastq.gz`)
```


# Create Raw ASV Count Table 

```{r raw-ASV-count-table}
# Raw ASV
raw_ASV_table <- makeSequenceTable(merged_ASVs)

# Intuition Check: Type and dimensions of the data
dim(raw_ASV_table)
typeof(raw_ASV_table)
class(raw_ASV_table)

# write out raw_asv_table 
write.table(raw_ASV_table, file = "data/01_DADA2/raw_ASV_counts.tsv",
            sep = "\t", quote = FALSE, col.names = NA)
```

# Assess ASV Quality 

## INTERPRETATION #1: What are the current ASV lengths? 

1. *What is the sequencing strategy?*
    - Illumina MiSeq paired-end 250bp (2x250bp) was used. 
    - 341F and 805R primers were used to amplify the V3-V4 hypervariable regions of the 16S rRNA gene.
2. *What's the total length of our starting amplicons?* 
    - Our primers are named after the location they bind to on the 16S gene. So: 
    - `805 - 341 = 464` 
    - The total length of the starting amplicon is 464 base pairs, with primers.
3. *What is the ASV length without primers?* 
    - The 341F (5′-CCTACGGGNGGCWGCAG) primer is 17 base pairs. 
    - The 805R (5′-GACTACHVGGGTATCTAATCC) primer is 21 base pairs. 
    - `464 - 17 - 21 = 426` 
    - The expected ASV length without primers is **426 base pairs**. 
    
4. *What is the length of the trimmed ASVs after `filterAndTrim()`?* 
    - In my `01_QualityTrimming.Rmd`, I used `trimLeft = c(17,21)` to remove the 17bp primer at the beginning of the forward reads and the 21bp primer at the beginning of the reverse reads. I also used `truncLen = c(249, 240)`, which removed 1 base at the end of the forward reads and removed 10 bases at the end of the reverse read. Therefore, the length of the trimmed ASV should be 426 base pairs, as the bases removed at the end of the read are just removed from the overlapping region of the ASV.
    
5. *What is the overlap between our forward and reverse reads?*
    - We performed 2x250 paired-end Illumina MiSeq Sequencing. Originally the overlapping region between forard and reverse reads was calculated to be 36bp (250+250-464=36). By trimming low quality bases from the end of our reads (one base from forward read and ten bases from reverse read) the overlapping region is calculated to be 25bp.


### ASV Length Stats 

```{r assess-ASV-length}
# Calculate summary stats
# Longest ASV?
maxLength_ASV <- max(nchar(getSequences(raw_ASV_table)))
maxLength_ASV

# Shortest ASV?
minLength_ASV <- min(nchar(getSequences(raw_ASV_table))) 
minLength_ASV

# Mean ASV length?
meanLength_ASV <- mean(nchar(getSequences(raw_ASV_table))) 
meanLength_ASV

# Median ASV length?
medianLength_ASV <- median(nchar(getSequences(raw_ASV_table))) 
medianLength_ASV

# Create a table to Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table)))
```
### ASV Length Plot 

Here's some of the stats that we've calculated:

- Max ASV length: `r maxLength_ASV` base pairs
- Min ASV length: `r minLength_ASV` base pairs
- Mean ASV length: `r meanLength_ASV` base pairs
- Median ASV length: `r medianLength_ASV` base pairs

Now, we will plot these ASV lengths to give us a better idea: 

```{r ASV-lengths-plot}
# Inspect the distribution of sequence lengths of all ASVs in data set 
# AFTER TRIM
plot_ASVLength_raw <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Raw ASV Length (bps)")

# Show the plot
plot_ASVLength_raw
```

INTERPRETATION #2: 

The ASV length data does not quite match my hypotheses from Interpretation #1. I expected most of the ASVs to have a length of 426bp. From the table we see that we have quite the variation in ASV length. From the plot we can visualize the variation, with most ASVs having a length between 390bp - 440bp. 

My shortest ASV length is 232bp, my mean ASV length is 391bp, and my median ASV length is 408bp. There is an 18bp difference between my expected median ASV length (426bp) and my actual median ASV length (408bp).


INTERPRETATION #3: Taking into account INTERPRETATIONS #1 and #2, what is your suggested trimming procedure for your ASV lengths? Will you allow some variation or trim to an exact length, similar to our in class example here?

Since I have quite a bit of variation in ASV length, yes I think I think I will allow some variation instead of trimming to an exact length. Based on the ASV length data I've generated above, I think I will trim ASV length to a range of 408bp - 429bp.Note that I was originally going to have this range be between 408 - 426bp (my median length to expected length), but extended the range to 429bp to include the large number of ASVs that were between 427 and 429bp in length. My original range of 408bp - 426bp only retained 18% of my ASVs.

## Trim ASV lengths 

Now that we are pretty convinced that we have some spurious ASVs, we will trim ASVs that are higher or lower than 408bp - 429bp base pairs in this data set. 


*Let's trim the ASVs to only be the right size, which is 408 - 426!* 

```{r trim-ASVs}
# Subset only ASVs that are between 408 and 426 bps long 
raw_ASV_table_trimmed <- 
  raw_ASV_table[,nchar(colnames(raw_ASV_table)) %in% 408:429]


# Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table_trimmed)))

# What proportion of total ASV sequences are left in the data? 
percRetained_Trimmed <- sum(raw_ASV_table_trimmed)/sum(raw_ASV_table)
percRetained_Trimmed # Show it 

# Inspect the distribution of sequence lengths of all ASVs in dataset 
# AFTER TRIM
plot_ASVLength_trimmed <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table_trimmed))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Trimmed ASV Length (bps)")

# Show the plot 
plot_ASVLength_trimmed
```

# Remove Chimeras


```{r rm-chimeras, fig.width=3.5, fig.height=3}
# Remove the chimeras in the raw ASV table
noChimeras_ASV_table <- 
  removeBimeraDenovo(raw_ASV_table_trimmed, 
                     method="consensus", 
                     multithread = n_threads, 
                     verbose=TRUE)

# Check the dimensions
dim(noChimeras_ASV_table)

# What proportion is left of the sequences? 
# Chimera removal compared to trimming 
percRetained_chimerasTrimmed <- sum(noChimeras_ASV_table)/sum(raw_ASV_table_trimmed)
percRetained_chimerasTrimmed
# Chimera removal compared to raw  
percRetained_chimerasRaw <-sum(noChimeras_ASV_table)/sum(raw_ASV_table)
percRetained_chimerasRaw

# Plot it 
plot_ASVLength_NoChimeras <- 
  data.frame(Seq_Length_NoChim = nchar(getSequences(noChimeras_ASV_table))) %>%
  ggplot(aes(x = Seq_Length_NoChim )) + 
  geom_histogram()+ 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs \n (Post-Chimera Removal)", 
       x = "ASV Length (bps)")

# Show the plot
plot_ASVLength_NoChimeras 
```

So, we have removed 47670 ASVs, which were chimeras. Let's review what proportion of ASVs we have retained along the workflow: 

1. `r round(percRetained_Trimmed * 100, digits = 2)`% of the raw ASVs were retained *after trimming*.
2. `r round(percRetained_chimerasTrimmed *100, digits = 2)`% of the trimmed ASV counts were retained *after chimera removal*.
3. This translates to `r round(percRetained_chimerasRaw *100, digits=2)`% retention of the original, raw merged ASV counts after *both trimming and chimera removal.*


INTERPRETATION #4: 

(1) What proportion of raw ASVs were retained after trimming?

- 66.83% of raw ASVs were retained after trimming?

(2) What proportion of the trimmed ASVs were retained after chimera removal?

- 92.06% of the trimmed ASVs were retained after chimera removal

(3) What proportion of the total merged ASVs were retained after both trimming and chimera removal? Do you think that the ASV lengths have been trimmed appropriately? For inspiration, you are encouraged to describe the conclusions from the the ASV length plot below.

- 61.52% of the total merged ASVs were retained after both trimming and chimera removal

At this point in time, I am confident that the ASV lengths have been trimmed appropriately. Initially in the ASV length trimming step, I established a length range of 408bp - 426bp and only 18.22% of the ASVs were retained! I feel much more confident in the range I used for acceptable ASV length, as I now retain 61.52% of the total merged ASVs were after BOTH trimming and chimera removal.

With my initial ASV length trimming parameters keeping ASVs that were between 408 and 426bp, the max number of ASVs at a certain length were just above 20 000. In the plot below you can see that now this number sits above 100 000. Even with chimera removal, the max number of ASVs after trimming is above 60 000. After a lot of troubleshooting, I am happy with these parameters and confident with the results I am getting.

Big thank you to Sophia for helping me figure this out!

### Plot ASV Lengths 

```{r plot-ASVLengths, fig.height=2.5, fig.width=7}
plot_ASVLength_raw + plot_ASVLength_trimmed + plot_ASVLength_NoChimeras + 
    plot_annotation(tag_levels = 'A')
```

# Track the read counts

Here, we will look at the number of reads that were lost in the filtering, denoising, merging, and chimera removal. 

```{r track-reads, fig.width=6, fig.height=4}
# A little function to identify number seqs 
getN <- function(x) sum(getUniques(x))

# Make the table to track the seqs 
track <- cbind(sapply(dada_forward, getN),
               sapply(dada_reverse, getN),
               sapply(merged_ASVs, getN),
               rowSums(noChimeras_ASV_table))

head(track)

# Update column names to be more informative (most are missing at the moment!)
colnames(track) <- c("denoisedF", "denoisedR", "merged", "nochim")
rownames(track) <- row.names(noChimeras_ASV_table)

# Generate a dataframe to track the reads through our DADA2 pipeline
track_counts_df <- 
  track %>%
  # make it a dataframe
  as.data.frame() %>%
  rownames_to_column(var = "sample_names")

# Now let's add a column for the number of ASVs
# First, intuition check that the samples match 
stopifnot(track_counts_df$sample_names == row.names(noChimeras_ASV_table))

# Now, let's add a new column with the number of ASVs
track_counts_df <- 
  track_counts_df %>%
  mutate(num_ASVs = rowSums(noChimeras_ASV_table > 1))

# Visualize it in table format 
DT::datatable(track_counts_df)

# Plot it!
track_counts_df %>%
  pivot_longer(denoisedF:nochim, names_to = "read_type", values_to = "num_reads") %>%
  mutate(read_type = fct_relevel(read_type, "denoisedF", "denoisedR", "merged", "nochim")) %>%
  ggplot(aes(x = read_type, y = num_reads, fill = read_type)) + 
  geom_line(aes(group = sample_names), color = "grey") + 
  geom_point(shape = 21, size = 3, alpha = 0.8) + 
  scale_fill_brewer(palette = "Spectral") + 
  labs(x = "Filtering Step", y = "Number of Sequences") + 
  theme_bw()
```
Now, let's plot the number of sequences that have been maintained in our samples using `geom_histogram()`. 

```{r numSeqsASV-plot, fig.height=2.5, fig.width=7}
plot_ReadDepth <- 
  track_counts_df %>%
  ggplot(aes(x = nochim)) + 
  geom_histogram() + 
  labs(x = "Total # of Sequences", y = "# of Samples") + 
  theme_bw()

# What is the ASV richness per sample? 
plot_ASVRichness <- 
  track_counts_df %>%
  ggplot(aes(x = num_ASVs)) + 
  geom_histogram() + 
  labs(x = "Total # of ASVs", y = "# of Samples") + 
  theme_bw()

# Now, let's look at the relationship of ASVs and Sequencing depth 
plot_ReadDepth_ASVRichness <- 
  track_counts_df %>%
  ggplot(aes(x = nochim, y = num_ASVs)) + 
  geom_point() + 
  labs(x = "Total # of Sequences", y = "# of ASVs") + 
  theme_bw()

# Show the plots together 
plot_ReadDepth + plot_ASVRichness + plot_ReadDepth_ASVRichness + 
    plot_annotation(tag_levels = 'A')
```

INTERPRETATION #5: What conclusions can you draw from each of the three individual plots above? Please note anything important that the figure might indicate regarding your project data. Are there any specific notes and considerations for later?

The figure above consists of three plots that summarize sequencing depth, ASV richness, and their relationship after DADA2 processing of 16S rRNA gene sequencing data.

**Panel A: Read Depth after DADA2:** Histogram of sequencing depth per sample (Total # of Sequences)

- The distribution is roughly unimodal, centered around 60 000 reads per sample.
- There are about 15 samples with very low read counts
- A couple of other samples have higher sequencing depth (>150,000 reads).

**Panel B: ASV Richness after DADA2:** Histogram of ASV richness (Total # of ASVs per sample)

- Most samples have ~250–1000 ASVs, following a similar unimodal distribution as sequencing depth.
- About 1 or 2 samples have very low ASV counts (~0-10). *Perhaps this corresponds to low read depth?*

**Panel C: Read Depth vs ASV Richness:** Scatter plot showing the relationship between sequencing depth and ASV richness.

- Similar to the class data, generally there might be a positive correlation, where samples with higher read depth tend to have more ASVs. 

**Overall Interpretation & Considerations**

1.	*Sequencing depth variation*: Most samples cluster around a common sequencing depth, but there are at least 15 samples that are likely under-sequenced.
2. *ASV richness follows read depth trends*: Samples with low sequencing depth exhibit reduced ASV richness, which may introduce bias in ecological interpretations
3.	*Rarefaction or normalization may be needed*: The plot on the right indicates that richness estimates are affected by read depth. As discussed in class, rarefying or using statistical corrections will be necessary for accurate comparative analysis.


# Assign Taxonomy 

```{r assign-tax}
# Assign up to genus level 
taxa_train <- 
  assignTaxonomy(noChimeras_ASV_table, 
                 refFasta = "/workdir/in_class_data/taxonomy/silva_nr99_v138.2_toGenus_trainset.fa.gz", 
                 multithread = n_threads)

# Add the genus/species information 
taxa_addSpecies <- 
  addSpecies(taxa_train, 
              refFasta = "/workdir/in_class_data/taxonomy/silva_v138.2_assignSpecies.fa.gz")

# Inspect the taxonomy 
glimpse(taxa_addSpecies) # Note that the rownames are the ASV sequences!
# Let's removing the ASV sequence rownames for display only
taxa_print <- taxa_addSpecies 
rownames(taxa_print) <- NULL
head(taxa_print)
#View(taxa_print)
```
# Export the Data

Remember from the beginning of the file under "Goals", we wanted to export 4 different types of information: 

1. ASV Count Table: `asv_table` (with and without sequence names)
2. ASV fasta file: `asvs_fasta` for building a phylogenetic tree at a later step.
3. Taxonomy Table  `tax_table`
4. Sample Information: `sample_data`  track the reads lots throughout DADA2 workflow. 


## 1. ASV Tables

We will export the following two ASV count tables, which will be in "long" format where the **ASVs are in rows** and the **sample names are in the columns.**

1. **With ASV seqs:** ASV headers include the *entire* ASV sequence.
2. **with ASV names:** This includes re-written and shortened headers like ASV_1, ASV_2, etc, which will match the names in our fasta file below.

### Structure of ASV tables

Let's check that the ASVs are in the rows and the sample names are in the columns. If not, we can use the transpose function `t()` in R. 

```{r structure-ASV-table}
# What's the current format of the ASV table?
head(rownames(noChimeras_ASV_table)) # Samples!
head(colnames(noChimeras_ASV_table)) # ASV Sequences

# Therefore, we need to transpose the matrix 
final_ASV_table_withSeqs <- t(noChimeras_ASV_table)

# Intuition check
head(rownames(final_ASV_table_withSeqs)) # ASV Sequences
head(colnames(final_ASV_table_withSeqs)) # Sample names
```

### Names in ASV tables

#### Fix Sample Names 

Notice above that the sample names are actually file names. Let's fix this! 


```{r SampleNames-ASV-table}
# Remember at the top of the file we created a vector of sample names 
head(sample_names)
# Let's check with the actual column names 
head(colnames(final_ASV_table_withSeqs)) # Sample names
# And then apply our sample name script to check, too
head(sapply(strsplit(colnames(final_ASV_table_withSeqs), "_"), `[`,1)) # Looks good! 

# Now, add a break in the script break if this isn't true! 
# Let's make sure the sample names match the file names in the matrix.
stopifnot(sapply(strsplit(colnames(final_ASV_table_withSeqs), "_"), `[`,1) == sample_names)

# Now, we've done some checks to prove to ourselves there will be no silent errors, 
# Let's rename! 
colnames(final_ASV_table_withSeqs) <- sample_names
head(colnames(final_ASV_table_withSeqs))
```

#### Rename ASVs

Then, we can also fix the names of our ASVs for our second ASV table where we will replace the ASV sequence names with ASV_1, ASV_2 etc..

```{r prepare-ASVcount-Seqtable}
# Give headers more manageable names
# First pull the ASV sequences from the rownames
ASV_seqs <- rownames(final_ASV_table_withSeqs)
ASV_seqs[1:5]

# How many ASVs? 
num_ASVs <- dim(final_ASV_table_withSeqs)[1] # select the number of rows
num_ASVs 

# Make an empty vector the length of the number of ASVs, 
# which is where we will place the new operational ASV names 
ASV_headers <- vector(num_ASVs, mode = "character")

# Let's mae sure we have an empty vector!
ASV_headers[1:5]
length(ASV_headers) # looks good! 

# Now, let's create a vector with ASV numbers
# loop through vector and fill it in with ASV names 
for (i in 1:num_ASVs) {
  # Add leading zero to ASV name so they print in correct order.
  ASV_number <- sprintf("%04d", i)
  # Now, rename each spot in the ASV header vector as we loop through the for loop
  ASV_headers[i] <- paste(">ASV", ASV_number, sep = "_")
}

# Intuition check
ASV_headers[1:5]

# Create a new ASV table, which will have the ASV numbers as names 
# View(noChimeras_ASV_table) # To view the table
final_ASV_table <- final_ASV_table_withSeqs
glimpse(final_ASV_table)

## Replace the ASV seqs with the ASV numbers 
row.names(final_ASV_table) <- sub(">", "", ASV_headers)
final_ASV_table[1:5, 1:5]
#View(final_ASV_table) # To view the table
``` 

### Write the ASV Tables!

1. **With ASV seqs:** ASV headers include the *entire* ASV sequence.
2. **with ASV names:** This includes re-written and shortened headers like ASV_1, ASV_2, etc, which will match the names in our fasta file below.

```{r write-asv-tables}
# 1. Write count table with ASV sequence names
write.table(final_ASV_table_withSeqs, 
            file = "data/01_DADA2/ASV_table_withSeqNames.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)

# 2. Write count table with ASV numbered names (e.g. ASV_1, ASV_2, etc)
write.table(final_ASV_table, 
            file = "data/01_DADA2/ASV_table.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)
```

## 2. ASV Fasta File 

Now, let's create a fasta file that has our ASV numbers with each of their corresponding ASV sequences. This file needs to be created because we will use it later to build a phylogenetic tree.

### Write the ASV Fasta File

```{r write-asv-fastas}
# Let's take our asv_headers
head(ASV_headers, n = 2)
# And combine it with their sequences
head(ASV_seqs, n = 2)

# Combine in a fasta format with the cbind() function
ASV_fasta <- c(rbind(ASV_headers, ASV_seqs))
head(ASV_fasta, n = 4)

# Then, let's write it to a fasta file!
# This will be our reference later on for which seq matches which ASV
write(ASV_fasta, "data/01_DADA2/ASVs.fasta")
```

## 3. Taxonomy Table 

Before we actually write the taxonomy table, let's put the ASV sequences as another column in our taxonomy table and then replace the rownames to be our numbered ASV names. 

### Reformat Taxonomy 

1. Move ASV sequences to column 
2. Replace ASV rownames with ASV numbered names


```{r reformat-tax-table}
# Inspect the taxonomy table
dim(taxa_addSpecies) # ASVs are in rows and Kingdom, Phylum, etc in Columns 
colnames(taxa_addSpecies) # Column names are Linnean Taxonomy 
head(rownames(taxa_addSpecies), n = 2) # ASV names are rownames 
class(taxa_addSpecies) # Character matrix

##### Prepare tax table 
# 1. Add the ASV sequences from the rownames to a column 
new_tax_table <- 
  taxa_addSpecies%>%
  as.data.frame() %>%
  rownames_to_column(var = "ASVseqs") 

# Intuition check 
glimpse(new_tax_table)

# IMPORTANT! Let's do our intuition check 
# This is where we ensure we don't mix up the ASV names!
stopifnot(new_tax_table$ASVseqs == rownames(final_ASV_table_withSeqs))

# Now let's add the ASV names 
rownames(new_tax_table) <- rownames(final_ASV_table)
head(new_tax_table)

### Final prep of tax table. Add new column with ASV names 
ASV_tax_table <- 
  new_tax_table %>%
  # add rownames from count table for phyloseq handoff
  mutate(ASV = rownames(final_ASV_table)) %>%
  # Reorder the columns
  dplyr::select(Kingdom, Phylum, Class, Order, Family, Genus, Species, ASV, ASVseqs)

# Assign the rownames, which is required by phyloseq
rownames(ASV_tax_table) <- ASV_tax_table$ASV

# Take a quick look
glimpse(ASV_tax_table)

# Intution check
stopifnot(ASV_tax_table$ASV == rownames(ASV_tax_table), 
          rownames(ASV_tax_table) == rownames(ASV_tax_table))
```

### Write the Taxonomy Table 

Now, let's write out the taxonomy table 

```{r write-tax-table}
# Write the table 
write.table(ASV_tax_table, 
            file = "data/01_DADA2/ASV_taxonomy.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)
```


## 4. Sample Data 

Let's save the `track_counts_df`, which can be useful for downstream analyses. 

```{r save-track reads}
# And save the track_counts_df a R object, which we will merge with metadata information in the next step of the analysis in nalysis/02_Taxonomic_Assignment. 
save(track_counts_df, file = "data/01_DADA2/track_read_counts.RData")
```

In this document I assigned ASVs on both forward and reverse reads by applying the error model, and then merged forward and reverse ASVs into "contiguous ASVs". I did quality trimming of ASV lengths, removed chimeras and assigned taxonomy with the Silva Database.

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```
