---
title: "PreProcessing ASVs with Phyloseq"
author: "Kailyn Hanke"
date: "2025-03-30"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center", 
                      # write figures to the figures folder
                      fig.path = "../figures/03_PreProcessing/")
```

# Goals 

First, we will use the [phyloseq package](https://joey711.github.io/phyloseq/) to combine all of our data objects that we exported from the DADA2 workflow (`asv_table`, `tax_table`, and metadata) and incorporate them into a single [specialized S4 R Data object](https://adv-r.hadley.nz/s4.html), known as a phyloseq object. Then, we will remove any potential contaminants and evaluate the accuracy of our sequencing run. Finally, we will write our our single `raw_preprocessed_physeq` phyloseq data object. 

## Specific Steps: 

1. Load in data that we've generated in `analysis/02_AssignASVs.Rmd` and fix all of the names to match each other. *The names must match for us to incorporate them into the S4 phyloseq object*: 
    a. `asv_table`: ASVs (rows) x Samples (columns)
    b. `tax_table`: ASV (rows) x Taxonomy (columns)
    c. `metadata`: Samples (rows) x All our data (*e.g. pH, Temp, treatment group, etc*; columns)
2. Combine the data into a phyloseq object. 
3. Remove any contaminating ASVs that are **chloroplasts**. 
4. Remove ASVs that are **mitochondria**. 
5. Evaluate any ASVs from the **negative controls**. Then, remove negative controls. 
6. Evaluate the mock community or **positive control** to learn the accuracy of sequencing. 
7. Check for **reverse complements**. 
8. Check the **sequencing depth** of samples. Remove samples that obviously have too few reads. 
9. Write a `raw_preprocessed_physeq` to be used in the next step of our workflow. 

## Input 

1. **Metadata**: `metadata.csv` and `data/01_DADA2/track_read_counts.RData`.
2. **ASV table**: `data/01_DADA2/ASV_table.csv` 
3. **Taxonomy Table**: `data/01_DADA2/ASV_taxonomy.tsv`

## Output 

1. A **pre-processed S4 phyloseq object**: `raw_preprocessed_physeq.RData`.



# Set Environment 

### Set the seed

```{r set-seed}
# Any number can be chosen
set.seed(111111)
```

## Load Packages 
```{r load-packages}
#install.packages("BiocManager")
#BiocManager::install("Biostrings")

# NOTE, you will need to install the following 2 packages to 
# re-create the negative control plots below!
#install.packages("ggpubr")
#install.packages("rstatix")

# Load packages with pacman
# Be sure that you load dada2 & Patchwork, which we didn't do in class on March 12th
pacman::p_load(devtools, phyloseq, dada2, patchwork, Biostrings, tidyverse,
               # Note to add ggpubr & rstatix!!
               ggpubr, rstatix, install = FALSE)
```

## Timing of this script

Let's record how long this file took to run on the class server, which we will record at the end of the script. 

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

Here, we will load in our **metadata** files, which include: 

1. `data/metadata.csv`: This file contains all of our samples and also any measured variables, including station, date of collection, depth of sample, temperature, pH, salinity, etc. 
2. `data/01_DADA2/track_read_counts.RData`: This file contains how many reads we maintained in our samples through the DADA2 workflow. 


```{r load-metadata}
# load in metadata
metadata_df <- 
  read_csv("data/metadata_amplicon_project.csv") %>%
  # Fix Column Name
  dplyr::rename("sample_names" = "Run") %>%
  # Fix additional column name (so can plot metadata)
  dplyr::rename("environment_material" = "environment_(material)") %>%
  dplyr::rename("environment_feature" = "environment_(feature)") %>%
  dplyr::rename("environment_biome" = "environment_(biome)") %>%
  # Add sample names also as a column 
  mutate(names = sample_names)


# Inspect 
head(metadata_df)
dim(metadata_df)

# include dada2 output
load("data/01_DADA2/track_read_counts.RData")

# Take a look
glimpse(track_counts_df)
dim(track_counts_df)

# Check filenames 
head(track_counts_df$sample_names)

# Fix sample names in track_reads 
track_counts_df$sample_names <- sapply(strsplit(track_counts_df$sample_names, "_"), `[`, 1)

# Intuition check 
head(track_counts_df$sample_names)

# What's different? 
setdiff(track_counts_df$sample_names, metadata_df$sample_names)

# Let's do a filtering join with left_join 
metadata_final_df <- 
  metadata_df %>%
  left_join(., track_counts_df, by = "sample_names") %>%
  # sample names to the rownames to merge into phyloseq
  column_to_rownames(var = "sample_names")

# Check 
dim(metadata_final_df)
```
### 1b. ASV Table 

Now, let's load in the ASV count table that we created within DADA2 in `analysis/02_AssignASVs.Rmd`. 

```{r load-asv-table}
asv_df <- 
  read.delim(file = "data/01_DADA2/ASV_table.tsv", sep = "\t",
           # add the column names and row names 
           header = TRUE, row.names = 1) 

# Final inspection 
head(colnames(asv_df))
asv_df[1:3, 1:3]
```
### 1c. Taxonomy Table

And, let's also load the taxonomy table that we created in `analysis/02_AssignASVs.Rmd`.

```{r load-tax-table}
tax_df <- 
  read.delim("data/01_DADA2/ASV_taxonomy.tsv", sep = "\t",
           header = TRUE, row.names = 1) 

# Inspect
dim(tax_df)
dim(asv_df)

# Add check 
stopifnot(rownames(asv_df) == rownames(tax_df))
```

# 2. Handoff to phyloseq 

This is where we will combine our metadata, asv count table, and taxonomy table into a single data S4 data object in R, which we will call `raw_physeq` for now.

```{r phyloseq-handoff}
raw_physeq <- 
  phyloseq(otu_table(asv_df, taxa_are_rows = TRUE),
         sample_data(metadata_final_df),
         tax_table(as.matrix(tax_df)))

# Check out 
raw_physeq

# save 
save(raw_physeq, file = "data/03_PreProcessing/raw_physeq.RData")
```

## 3. Remove chloroplasts

```{r rm-chloro}
noChloros_physeq <- 
  raw_physeq %>%
  subset_taxa(Order != "Chloroplast" | is.na(Order))

# How many ASVs were chloroplasts? 
numChloros_ASVs <- ntaxa(raw_physeq) - ntaxa(noChloros_physeq)
```

There were `r numChloros_ASVs` ASVs that were chloroplasts.

## 4. Remove Mitochondria 

```{r rm-mitos}
noChlorosMitos_physeq <-  
  noChloros_physeq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family)) 

# How many ASVs were mitochondria? 
numMitos_ASVs <- ntaxa(noChloros_physeq) - ntaxa(noChlorosMitos_physeq)

```

There were `r numMitos_ASVs` ASVs that were mitochondria in the data set.


Interpretation #1: How many chloroplast and/or mitochondrial ASVs in your dataset? Does this seem reasonable?

There were 36 chloroplast ASVs and 450 mitochondrial ASVs identified in my dataset. I was surprised that only 36 chloroplast ASVs were identified due to the large number of samples and how deeply sequenced this dataset is. Even though 450 mitochondrial ASVs seems like quite a large number, I think this is reasonable, given that in my dataset there are over 75 000 ASVs identified.


# Evaluate and remove the control samples 

Now, let's take a look at the negative controls and then make a decision about whether or not to remove the ASVs that we found in our controls 

1. Negative controls 
2. ASVs found within the negative controls and their distribution in the samples. 
3. Evaluate the mock community 

## 5. Negative Controls 

INTERPRETATION #2: Did your dataset have any negative controls? If so, how many ASVs did you remove? Does that seem reasonable to you?

No, my dataset did not have any negative controls! Because of this I did not remove any ASVs. Like Mar stated, negative controls are not optional, so it doesn't seem very reasonable to me that there were no negative controls in this dataset.

## 6. Positive Controls


INTERPRETATION #3: Did your dataset have a mock community? If so, what can you say about the accuracy of your sequencing run?

My dataset did not have a mock community, so I was not able to make any conclusions about the accuracy of my sequencing run.


# 7. Reverse Complements


```{r reverse-complements}
# Pull out Taxa Table
tax_table <- 
  noChlorosMitos_physeq %>%
  tax_table() %>% 
  data.frame()

# Grab ASV Sequences and Names
asv_names <- tax_table$ASV
asv_seqs <- tax_table$ASVseqs

# Create empty comparison matrix
asv_rev_comp <- matrix(nrow=length(asv_seqs), ncol = length(asv_seqs)) 

# Fix the names in the rows and columns 
rownames(asv_rev_comp) <- asv_names
colnames(asv_rev_comp) <- asv_names

# Convert sequences to Biostrings
asv_dna <- DNAStringSet(asv_seqs) 

# Inspect 
head(asv_dna)

# Construct reverse complement
asv_rev_dna <- reverseComplement(asv_dna) 

# Now loop through every asv sequence to check 
for(i in 1:length(asv_seqs)){ # For each asv sequence...
  match_vec <- asv_dna[i] == asv_rev_dna # Compare it to the reverse complement of every other sequence...
  asv_rev_comp[,i] <- match_vec # Write that logical vector as a new column 
}

# Find how many TRUEs (matches) we have, divide by 2 because each pair occurs twice
cat("For", sum(asv_rev_comp) / 2,"ASVs, the reverse complement will need to be removed") 
```

INTERPRETATION #4: Do you have any reverse complements in your dataset?

No I do not have any reverse complements that need to be removed.

# 8. Sequencing Depth

Let's take a look! 

```{r seq-depth, fig.height=3, fig.width=8}
# The current data objec
noChlorosMitos_physeq

# What is the library size/sequencing depth for each sample? 
seqSums_df <- 
  noChlorosMitos_physeq %>%
  otu_table() %>%
  # Sum each sample column 
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "names") %>%
  left_join(., metadata_final_df, by = "names") 

# Rename second column 
colnames(seqSums_df)[2] <- "TotalSeqs"

# check
dim(seqSums_df)
head(seqSums_df)

# Show the depth of samples 
seqSums_df %>%
  dplyr::select(names, TotalSeqs) %>%
  arrange(TotalSeqs) %>%
  head()

# plot it as a bar plot 
numSeq_bar_plot <- 
  seqSums_df %>%
  ggplot(aes(x=reorder(names, TotalSeqs), y = TotalSeqs,
             fill = environment_material)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_blank()) + 
  labs(y = "Read Depth", x = "Sample") + 
  theme(legend.position = "none")

# histogram
numSeq_hist_plot <- 
  seqSums_df %>%
  ggplot(aes(x= TotalSeqs, fill = environment_material)) + 
  geom_histogram(color = "black") + 
  labs(y = "# of Samples", x = "Read Depth") + 
  theme(legend.position = "bottom")

# Density plot 
numSeq_density_plot <- 
  seqSums_df %>%
  ggplot(aes(TotalSeqs, fill = environment_material)) +
  geom_density(alpha = 0.5) + 
  labs(x = "Read Depth") + 
  theme(legend.position = "none")

# Put it all together 
numSeq_bar_plot + numSeq_hist_plot + numSeq_density_plot + 
  plot_annotation(tag_levels = "A") 
```


INTERPRETATION #5: What can you conclude about your sequencing depths? What are your sample distributions? Do you have "enough" reads? Are there any samples that should be removed at this step?


From these plots I can conclude that the majority of my samples have a read depth between (approximately) 45000 and 100000 which is great. Something else I immediately noticed from these plots is that it appears that 15 samples have a read depth that is very low. In the code below I checked to see what the minimum sequencing depth is, and I discovered it was 533. I then made a dataframe to look at which samples had low sequencing depth. I decided to remove all the samples with a read depth below 1500, like Mar did in their example. To do this, I removed seven samples: ERR3143462, ERR3143466, ERR3143470, ERR3143474, ERR3143460, ERR3143465, ERR3143464. After doing this processing the minimum read depth for a sample is 1524. Will confirm with Mar/Sophia if this is an appropriate cutoff for my dataset.



```{r rm-samps}
# What's the min seq depth? 

min(sample_sums(noChlorosMitos_physeq))


#Making a dataframe to inspect further.
sample_depths <- data.frame(
  Sample = sample_names(noChlorosMitos_physeq),
  Depth = sample_sums(noChlorosMitos_physeq)
)

# Sort it from lowest to highest
sample_depths_sorted <- sample_depths[order(sample_depths$Depth), ]

head(sample_depths_sorted, n=20)  # View lowest-depth samples

# Remove the seven samples
raw_preprocessed_physeq <- 
  noChlorosMitos_physeq %>%
  subset_samples(!(names %in% c("ERR3143462", "ERR3143466", "ERR3143470", 
                                "ERR3143474", "ERR3143460", "ERR3143465", "ERR3143464")))

min(sample_sums(raw_preprocessed_physeq))
```

# 9. Save Output 

## Raw Preprocessed Phyloseq Object

Finally, let's save our phyloseq object, which we will use for the next step! 

```{r save-physeq}
save(raw_preprocessed_physeq, file = "data/03_PreProcessing/raw_preprocessed_physeq.RData")
```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```
