---
title: "Between-Sample (Beta) Diversity of microbes in dairy samples across seasons"
author: "Kailyn Hanke"
date: "2025-04-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/09_Seasonal_Ordination/")
```

#Goals

The goal of this document is to run an analysis on the between-sample (beta( diversity of the samples in my project, similar to the analysis that I ran in 06_Ordination. This is the second of three new analysis documents I will make (I have previously made 08_Seasonal_Biodiversity and will subsequently make 10_Seasonality_Composition). See more info on my organizational layout under "Goals" in 08_Seasonal_Biodiversity.

Unlike in my original 06_Ordination analysis, I do not need to evaluate sequencing depth and normalize read counts, as I have previously done this in my original homework document!

In particular, I'm hoping that this seasonal/sample analysis will explain the weird grouping in my Skimmed Milk Powder samples.

1. Calculate community **dis**similarities. Numbers between 0 and 1. If 0, completely similar versus if they are 1, then they're completely dissimilar. Here I will split up my communities by sample type *AND* season!!
    a. **Sorensen**: Shared Species as a binary value: Abundance-unweighted 
    b. **Bray-Curtis**: Shared Abundant species: Abundance-weighted
    c. **(Abundance-)Weighted UNIFRAC**: Consider Abundant Species and where they fall on the tree  
5. Run statistics with functions from the vegan R package: 
    a. PERMANOVA with `adonis2()`.
    b. betadispR with `betadisper()` and `permutest()`. 
6. Visualize the community data with two unconstrained Ordinations:  
    a. **PCoA**: Linear Method. Eigenvalue = how much variation is explained by each axis. Choose to view axis 1, 2, 3, etc. and plot them together.  
    b. **NMDS**: Non-linear. Smush multiple Dimensions into 2 or 3 axes. Need to report Stress value (ideally <0.15).
    

## Inputs
    
1. We will need the `scaled_physeq.RData`, that we created in `analysis/06_Ordination.Rmd`. 

## Outputs

While defining community based on sample type and season:
1. Calculated beta-diversity dissimilarity measures (*e.g.* Sorensen, Bray-Curtis, abundance-unweighted and -weighted UniFrac) across every sample. 
2. Ordination figures (*i.e.,* PCoA/NMDS) to include in the scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests (*i.e.,* PERMANOVA & betadisper) conveying the measured and quantified changes and patterns in biodiversity.

# Set up 

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed

```{r set-seed}
set.seed(111111)
```

## Load Packages, colors & functions  
```{r load-packages}
pacman::p_load(tidyverse, devtools, phyloseq, patchwork, vegan, 
               install = FALSE)

# Load Colors 
source("code/colors.R")

# Load functions 
source("code/functions.R")


```


# 1. Load in the data 

```{r load-data}

load("data/06_Ordination/scaled_physeq.RData")

read.csv("data/06_Ordination/metadata_df_cut20000.csv")

load("data/06_Ordination/metadata_df_cut20000.RData")

```

# Beta Diversity: Dissimilarity Metrics in Microbial Ecology

# Calculating dissimilarity

Now, let's make distance objects of the dissimiliarity measures we used in class!

```{r calc-dissimilarity}

# Sorensen Dissimiliarty
scaled_sorensen_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = TRUE)

# Bray-Curtis Dissimiliarty
scaled_bray_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = FALSE)

# Abundance-Unweighted UniFrac
scaled_uUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "unifrac")

# Abundance-Weighted UniFrac
scaled_wUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "wunifrac")

```


# PERMANOVA: Testing Means/Centroids

Here, we will focus on comparing the Sorensen and the Bray-Curtis. 

First I will mutate my metadata to include a new column season to define mid- and late-lactation.


```{r mutate-add-season-df}

# Create a new column 'season' based on the Collection_Date
metadata_df_season <- metadata_df_cut20000 %>%
  mutate(season = case_when(
    month(as.Date(Collection_Date, format = "%m/%d/%Y")) == 5 ~ "mid-lactation",  # May is mid-lactation
    TRUE ~ "late-lactation"  # All other months are late-lactation
  ))

# Check if the new column is created correctly
table(metadata_df_season$season)

glimpse(metadata_df_season)

#Save
save(metadata_df_season, 
     file = "data/09_Seasonal_Ordination/metadata_df_season.RData")

```

### Sorensen

```{r PERMANOVA-categorical-sorensen}
# Sorensen
## 1. Run with by = terms for R² values, sensitive to order of variables! 
## ALWAYS check and confirm the order of your terms and how they interact with each other.  
sorensen_sample_adonis_terms1 <- adonis2(scaled_sorensen_dist ~ environment_material * season, data = metadata_df_season, by = "terms")

sorensen_sample_adonis_terms1

# Check the order of the terms
sorensen_sample_adonis_terms2 <- adonis2(scaled_sorensen_dist ~ season * environment_material, data = metadata_df_season, by = "terms")

sorensen_sample_adonis_terms2

## 2. Run with by = "margin" for marginal p-values, which we can compare to the residuals from the first one. 

sorensen_sample_adonis_margin <- adonis2(scaled_sorensen_dist ~ season * environment_material, data = metadata_df_season, by = "margin")

sorensen_sample_adonis_margin
```

### Bray-Curtis

```{r PERMANOVA-categorical-bray}
# Bray-Curtis
## ALWAYS check and confirm the order of your terms and how they interact with each other.  
bray_sample_adonis_terms1 <- adonis2(scaled_bray_dist ~ environment_material * season, data = metadata_df_season, by = "terms")

bray_sample_adonis_terms1

bray_sample_adonis_terms2 <- adonis2(scaled_bray_dist ~ season * environment_material, data = metadata_df_season, by = "terms")

bray_sample_adonis_terms2


## 2. Run with by = "margin" for marginal p-values
bray_sample_adonis_margin <- adonis2(scaled_bray_dist ~ season * environment_material, data = metadata_df_season, by = "margin")

bray_sample_adonis_margin
```


### Unweighted-Unifrac

```{r PERMANOVA-categorical-uwunifrac}
# Unweighted Unifrac
## ALWAYS check and confirm the order of your terms and how they interact with each other.
uUnifrac_sample_adonis_terms1 <- adonis2(scaled_uUnifrac_dist~ environment_material * season, data = metadata_df_season, by = "terms")

uUnifrac_sample_adonis_terms1

uUnifrac_sample_adonis_terms2 <- adonis2(scaled_uUnifrac_dist ~ season * environment_material, data = metadata_df_season, by = "terms")

uUnifrac_sample_adonis_terms2

## 2. Run with by = "margin" for marginal p-values
uUnifrac_sample_adonis_margin <- adonis2(scaled_uUnifrac_dist ~ season * environment_material, data = metadata_df_season, by = "margin")

uUnifrac_sample_adonis_margin
```


### Weighted-Unifrac

```{r PERMANOVA-categorical-wunifrac}
# Weighted Unifrac
## ALWAYS check and confirm the order of your terms and how they interact with each other.
wUnifrac_sample_adonis_terms1 <- adonis2(scaled_wUnifrac_dist~ environment_material * season, data = metadata_df_season, by = "terms")

wUnifrac_sample_adonis_terms1

wUnifrac_sample_adonis_terms2 <- adonis2(scaled_wUnifrac_dist ~ season * environment_material, data = metadata_df_season, by = "terms")

wUnifrac_sample_adonis_terms2

## 2. Run with by = "margin" for marginal p-values
wUnifrac_sample_adonis_margin <- adonis2(scaled_wUnifrac_dist ~ season * environment_material, data = metadata_df_season, by = "margin")

wUnifrac_sample_adonis_margin
```

# Betadisper: Testing Variances/Dispersions

## Sorensen

```{r betadisper-sorensen}

# Homogeneity of Disperson test with beta dispr

# Sorensen Betadisper - Sample Type 
dispr_sorensen_sample <- betadisper(scaled_sorensen_dist, metadata_df_season$environment_material)
# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).

permutest(dispr_sorensen_sample)

# Sorensen Betadisper - Season  
dispr_sorensen_season <- betadisper(scaled_sorensen_dist, metadata_df_season$season)

permutest(dispr_sorensen_season)

```

## Bray-Curtis

```{r betadisper-bc}

# Homogeneity of Disperson test with beta dispr

# Bray-Curtis Betadisper - Sample Type 
dispr_bray_sample <- betadisper(scaled_bray_dist, metadata_df_season$environment_material)
# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).

permutest(dispr_bray_sample)

# Bray-Curtis Betadisper - Season  
dispr_bray_season <- betadisper(scaled_bray_dist, metadata_df_season$season)

permutest(dispr_bray_season)

```

## Unweighted-Unifrac

```{r betadisper-uunifrac}

# Homogeneity of Disperson test with beta dispr

# Unweighted Unifrac Betadisper - Sample Type 
dispr_uUnifrac_sample <- betadisper(scaled_uUnifrac_dist, metadata_df_season$environment_material)
# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).

permutest(dispr_uUnifrac_sample)

# Unweighted Unifrac Betadisper - Season  
dispr_uUnifrac_season <- betadisper(scaled_uUnifrac_dist, metadata_df_season$season)

permutest(dispr_bray_season)

```

## Weighted-Unifrac

```{r betadisper-wunifrac}

# Homogeneity of Disperson test with beta dispr

# Weighted Unifrac Betadisper - Sample Type 
dispr_wUnifrac_sample <- betadisper(scaled_wUnifrac_dist, metadata_df_season$environment_material)
# permutest() performs a non-parametric permutation test, which is robust and valid for the kind of data used in beta diversity analysis (e.g., dissimilarity matrices).

permutest(dispr_wUnifrac_sample)

# Weighted Unifrac Betadisper - Season  
dispr_wUnifrac_season <- betadisper(scaled_wUnifrac_dist, metadata_df_season$season)

permutest(dispr_wUnifrac_season)

```

# 6. Visualize Community Dissimilarity with Ordination

## PCoA: Principal Coordinates Analysis

```{r update-metadata-df}

#To include seasons, need to update the sample_data in my phyloseq object

sample_data(scaled_physeq) <- sample_data(metadata_df_season)

#Intuition check

head(sample_data(scaled_physeq))

scaled_physeq_season <- sample_data(scaled_physeq)

# Save the updated phyloseq object
save(scaled_physeq_season, file = "data/09_Seasonal_Ordination/scaled_physeq_season.RData")


```

### Sorensen

```{r pcoa-sorensen, fig.height=3.5, fig.width=7}

# First, calculate PCoA with Soresen
scaled_soren_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = TRUE)

# Take a quick look
str(scaled_soren_pcoa)

# Plot it: Sorensen PCoA  
sorensen_pcoa_plot <- 
  plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_pcoa,
                color = "environment_material",
                shape = "season",
                title = "Sorensen PCoA") + 
  scale_color_manual(values = environment_colors) + 
  scale_shape_manual(values = c(15, 16)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = environment_material)) + 
  theme_bw() + 
  theme(legend.position = "right")

sorensen_pcoa_plot

```

### Bray-Curtis

```{r pcoa-bray, fig.height=3.5, fig.width=7}

# First, calculate PCoA with Bray
scaled_bray_pcoa <- 
  ordinate(physeq = scaled_physeq,
         method = "PCoA",
         distance = "bray", binary = FALSE)

# Take a quick look
str(scaled_bray_pcoa)

# Plot it: Bray PCoA  
bray_pcoa_plot <- 
  plot_ordination(physeq = scaled_physeq,
                ordination = scaled_bray_pcoa,
                color = "environment_material",
                shape = "season",
                title = "Bray-Curtis PCoA") + 
  scale_color_manual(values = environment_colors) + 
  scale_shape_manual(values = c(15, 16)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = environment_material)) + 
  theme_bw() + 
  theme(legend.position = "right")

bray_pcoa_plot

```

### Unweighted-Unifrac

```{r pcoa-uwunifrac, fig.height=3.5, fig.width=7}

# First, calculate PCoA with unweighted UniFrac
scaled_uwunifrac_pcoa <- 
  ordinate(physeq = scaled_physeq,
           method = "PCoA",
           distance = "unifrac",  # change here!
           weighted = FALSE)      

# Take a quick look
str(scaled_uwunifrac_pcoa)

# Plot it: Unweighted UniFrac PCoA  
uwunifrac_pcoa_plot <- 
  plot_ordination(physeq = scaled_physeq,
                  ordination = scaled_uwunifrac_pcoa,
                  color = "environment_material",
                  shape = "season",
                  title = "Unweighted-UniFrac PCoA") + 
  scale_color_manual(values = environment_colors) + 
  scale_shape_manual(values = c(15, 16)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = environment_material)) + 
  theme_bw() + 
  theme(legend.position = "right")

uwunifrac_pcoa_plot


```

### Weighted-Unifrac

```{r pcoa-wunifrac, fig.height=3.5, fig.width=7}

# First, calculate PCoA with Weighted-UniFrac
scaled_wunifrac_pcoa <- 
  ordinate(physeq = scaled_physeq,
           method = "PCoA",
           distance = "unifrac",  
           weighted = TRUE)      

# Take a quick look
str(scaled_wunifrac_pcoa)

# Plot it: Weighted-UniFrac PCoA  
wunifrac_pcoa_plot <- 
  plot_ordination(physeq = scaled_physeq,
                  ordination = scaled_wunifrac_pcoa,
                  color = "environment_material",
                  shape = "season",
                  title = "Weighted-UniFrac PCoA") + 
  scale_color_manual(values = environment_colors) + 
  scale_shape_manual(values = c(15, 16)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = environment_material)) + 
  theme_bw() + 
  theme(legend.position = "right")

wunifrac_pcoa_plot

```

Now show all the PCoA plot together!

```{r combine-pcoa}

# Show the plots 
sorensen_pcoa_plot + bray_pcoa_plot + uwunifrac_pcoa_plot + wunifrac_pcoa_plot +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")

```

## NMDS: Non-Metric Multidimensional Scaling

### Sorensen

```{r nmds-sorensen, fig.height=3.5, fig.width=7}

## SORENSEN 
scaled_soren_nmds <- 
  ordinate(physeq = scaled_physeq,
         method = "NMDS",
         distance = "bray", binary = TRUE)

# Plot it! 
sorensen_nmds_plot <- 
  plot_ordination(physeq = scaled_physeq,
                ordination = scaled_soren_nmds,
                color = "environment_material",
                shape = "season",
                title = "Sorensen NMDS") + 
  scale_color_manual(values = environment_colors) + 
  scale_shape_manual(values = c(15, 16)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = season)) + 
  theme_bw() + 
  theme(legend.position = "right")

sorensen_nmds_plot

```

### Bray-Curtis

```{r nmds-bray, fig.height=3.5, fig.width=7}

#Calculate NMDS with Bray-Curtis
scaled_bray_nmds <- 
  ordinate(physeq = scaled_physeq,
         method = "NMDS",
         distance = "bray", binary = FALSE)

# Plot it: Bray-Curtis NMDS 
bray_nmds_plot <- 
  plot_ordination(physeq = scaled_physeq,
                ordination = scaled_bray_nmds,
                color = "environment_material",
                shape = "season",
                title = "Bray-Curtis NMDS") + 
  scale_color_manual(values = environment_colors) + 
  scale_shape_manual(values = c(15, 16)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = season)) + 
  theme_bw() + 
  theme(legend.position = "right")

bray_nmds_plot

```

### Unweighted-Unifrac

```{r nmds-unweighted-unifrac, fig.height=3.5, fig.width=7}

#Calculate NMDS with Unweighted-Unifrac
scaled_uwunifrac_nmds <- 
  ordinate(physeq = scaled_physeq,
         method = "NMDS",
         distance = "unifrac", weighted = FALSE)

# Plot it: Bray-Curtis NMDS 
uwunifrac_nmds_plot <- 
  plot_ordination(physeq = scaled_physeq,
                ordination = scaled_uwunifrac_nmds,
                color = "environment_material",
                shape = "season",
                title = "Unweighted-Unifrac NMDS") + 
  scale_color_manual(values = environment_colors) + 
  scale_shape_manual(values = c(15, 16)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = season)) + 
  theme_bw() + 
  theme(legend.position = "right")

uwunifrac_nmds_plot

```

### Weighted-Unifrac

```{r nmds-weighted-unifrac, fig.height=3.5, fig.width=7}

#Calculate NMDS with Weighted-Unifrac
scaled_wunifrac_nmds <- 
  ordinate(physeq = scaled_physeq,
         method = "NMDS",
         distance = "unifrac", weighted = TRUE)

# Plot it: Bray-Curtis NMDS 
wunifrac_nmds_plot <- 
  plot_ordination(physeq = scaled_physeq,
                ordination = scaled_wunifrac_nmds,
                color = "environment_material",
                shape = "season",
                title = "Weighted-Unifrac NMDS") + 
  scale_color_manual(values = environment_colors) + 
  scale_shape_manual(values = c(15, 16)) + 
  geom_point(size = 5, alpha = 0.5, aes(color = season)) + 
  theme_bw() + 
  theme(legend.position = "right")

wunifrac_nmds_plot

```

Now combine all the NMDS plots together!

```{r combine-nmds}

# Show the plots 
sorensen_nmds_plot + bray_nmds_plot + uwunifrac_nmds_plot + wunifrac_nmds_plot +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")

```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```
