---
title: "Between-Sample (Beta) Diversity of microbes in dairy samples across seasons"
author: "Kailyn Hanke"
date: "2025-04-26"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/09_Seasonal_Ordination/")
```

#Goals

The goal of this document is to run an analysis on the between-sample (beta( diversity of the samples in my project, similar to the analysis that I ran in 06_Ordination. This is the second of three new analysis documents I will make (I have previously made 08_Seasonal_Biodiversity and will subsequently make 10_Seasonality_Composition). See more info on my organizational layout under "Goals" in 08_Seasonal_Biodiversity.

Unlike in my original 06_Ordination analysis, I do not need to evaluate sequencing depth and normalize read counts, as I have previously done this in my original homework document!

1. Calculate community **dis**similarities. Numbers between 0 and 1. If 0, completely similar versus if they are 1, then they're completely dissimilar. Here I will split up my communities by sample type *AND* season!!
    a. **Sorensen**: Shared Species as a binary value: Abundance-unweighted 
    b. **Bray-Curtis**: Shared Abundant species: Abundance-weighted
    c. **(Abundance-)Weighted UNIFRAC**: Consider Abundant Species and where they fall on the tree  
5. Run statistics with functions from the vegan R package: 
    a. PERMANOVA with `adonis2()`.
    b. betadispR with `betadisper()` and `permutest()`. 
6. Visualize the community data with two unconstrained Ordinations:  
    a. **PCoA**: Linear Method. Eigenvalue = how much variation is explained by each axis. Choose to view axis 1, 2, 3, etc. and plot them together.  
    b. **NMDS**: Non-linear. Smush multiple Dimensions into 2 or 3 axes. Need to report Stress value (ideally <0.15).
    

## Inputs
    
1. We will need the `scaled_physeq.RData`, that we created in `analysis/06_Ordination.Rmd`. 

## Outputs

While defining community based on sample type and season:
1. Calculated beta-diversity dissimilarity measures (*e.g.* Sorensen, Bray-Curtis, abundance-unweighted and -weighted UniFrac) across every sample. 
2. Ordination figures (*i.e.,* PCoA/NMDS) to include in the scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests (*i.e.,* PERMANOVA & betadisper) conveying the measured and quantified changes and patterns in biodiversity.

# Set up 

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed

```{r set-seed}
set.seed(111111)
```

## Load Packages, colors & functions  
```{r load-packages}
pacman::p_load(tidyverse, devtools, phyloseq, patchwork, vegan, 
               install = FALSE)

# Load Colors 
source("code/colors.R")

# Load functions 
source("code/functions.R")


```


# 1. Load in the data 

```{r load-data}

load("data/06_Ordination/scaled_physeq.RData")

```

# Beta Diversity: Dissimilarity Metrics in Microbial Ecology

# 4. Calculating dissimilarity

Now, let's make distance objects of the dissimiliarity measures we used in class!

```{r calc-dissimilarity}

# Sorensen Dissimiliarty
scaled_sorensen_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = TRUE)

# Bray-Curtis Dissimiliarty
scaled_bray_dist <- phyloseq::distance(scaled_physeq, method = "bray", binary = FALSE)

# Abundance-Unweighted UniFrac
scaled_uUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "unifrac")

# Abundance-Weighted UniFrac
scaled_wUnifrac_dist <- phyloseq::distance(scaled_physeq, method = "wunifrac")

```



