---
title: "Microbial Compositional Analysis of Mid- and Late-lactation Dairy Samples
  Along a Processing Pipeline"
author: "Kailyn Hanke"
date: "2025-04-27"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      fig.path = "../figures/10_Composition/")
```

# Goals 

The goal of this document is to run an analysis of micorbial composition of mid- and late-lactation dairy samples across a processing pipeline, similar to the analysis that I ran in 07_Composition. This is the third of three new analysis documents I will make (I have previously made 08_Seasonal_Biodiversity and 09_Ordination). See more info on my organizational layout under "Goals" in 08_Seasonal_Biodiversity.

1. Load in scaled/normalized phyloseq object. 
2. Calculate the relative abundances of taxonomic groups at various levels: 
    A. Phylum
    B. Genus
    C. ASV 
3. Plot them, and narrow in on specific taxonomic group of interest


## Inputs

1. We will need the `scaled_physeq_season.RData`, which includes a rooted tree that we created in `analysis/09_Seasonal_Ordination/scaled_physeq_season.RData`.

## Outputs 

1. Beautiful visualizations of microbial taxa and how they vary across dairy sample types (including mid- and late-lactation information.
2. Run some stats! 

# Setup 

## Load Packages 
```{r load-packages}
# Load in the packages 
pacman::p_load(tidyverse, devtools, DT, phyloseq, patchwork, rstatix, 
               install = FALSE)

# load colors
source("code/colors.R")
```

## 1. Load in Scaled Phyloseq object

The following phyloseq object contains microbial community composition data in a standardized format. In this case, we’ve already normalized the reads (scaled to 20090 per sample and including season data), which is essential for comparing relative abundances.

```{r load-data}

load("data/09_Seasonal_Ordination/scaled_physeq_season.RData")

load("data/06_Ordination/scaled_physeq.RData")

load("data/09_Seasonal_Ordination/metadata_df_season.RData")

#Intuition check
min(sample_sums(scaled_physeq)) #19933

#Note for some reason the scaled_phyloseq_season wasn't loading in! Will create scaled_physeq_season again until I fix the error

```

First I will mutate my metadata to include a new column season to define mid- and late-lactation.


```{r update-physeq-df}

#To include seasons, need to update the sample_data in my phyloseq object

# Update the sample_data to include season
sample_data(scaled_physeq) <- sample_data(metadata_df_season)

# Save the updated phyloseq object
scaled_physeq_season <- scaled_physeq

```


# Taxonomic Analysis! 

## A. Phylum 

```{r calc-phylum-df}

phylum_df_season <- 
  scaled_physeq_season %>%
  # Agglomerate all ASV counts within a phylum
  tax_glom(taxrank = "Phylum") %>%
  # Calculate the relative abundance! 
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  # Create a dataframe from phyloseq object
  psmelt() %>%
  #Filter out Phyla < 1 % 
  #dplyr::filter(Abundance > 0.01) %>%
  #Fix the station order 
  mutate(environment_material = fct_relevel(environment_material, c(
    "Bulk Tank Milk", "Collection Tanker Milk", "Whole Milk Silo", "Cream Silo", "Skimmed Milk Silo", "Skimmed Milk Powder")))

#What are the phylum abundances?
phylum_df_season %>%
  group_by(Phylum, season) %>%
  summarize(mean_PercAbund = round(mean(Abundance), digits = 4), .groups = "drop") %>%
  arrange(season, -mean_PercAbund) %>%
  datatable()


top10_phyla_season <- 
  phylum_df_season %>%
  group_by(season, Phylum) %>%
  # group by BOTH season and phylum
  summarize(mean_PercAbund = mean(Abundance), .groups = "drop") %>%
  # regroup by season
  group_by(season) %>%               
  # take top 10 per season
  slice_max(order_by = mean_PercAbund, n = 10) %>%  
  arrange(season, -mean_PercAbund)  

#Identify top phyla in mid-lactation samples
top10_phyla_mid <- 
  phylum_df_season %>%
  filter(season == "mid-lactation") %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = mean(Abundance)) %>%
  arrange(-mean_PercAbund) %>%
  slice_head(n = 10) %>%
  pull(Phylum) #comment out to have mini df w/ abundances

#View
top10_phyla_mid

#Identify top phyla in late-lactation samples
top10_phyla_late <- 
  phylum_df_season %>%
  filter(season == "late-lactation") %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = mean(Abundance)) %>%
  arrange(-mean_PercAbund) %>%
  slice_head(n = 10) %>%
  pull(Phylum) #comment out to have mini df w/ abundances


#View
top10_phyla_late

```

## Stacked Bar plots 

### Mid-lactation
```{r phylum-stacked-bar-ml, fig.width=9, fig.height=3.5}

#Intuition check all phyla in top ten for mid-lactation in phylum_colors (in my previous composition doc they were not and it caused me a bunch of avoidable issues)

top10_phyla_mid %in% names(phylum_colors)


#actual plot (by sample)
stacked_plot_mid <- phylum_df_season %>%
  filter(Phylum %in% top10_phyla_mid) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Top 10 Mid-Lactation Phyla Across All Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))

stacked_plot_mid

```


### Late-lactation
```{r phylum-stacked-bar-ll, fig.width=9, fig.height=3.5}

#Intuition check all phyla in top ten for late-lactation in phylum_colors (in my previous composition doc they were not and it caused me a bunch of avoidable issues)

top10_phyla_late %in% names(phylum_colors)



#actual plot (by sample)
stacked_plot_late <- phylum_df_season %>%
  filter(Phylum %in% top10_phyla_late) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Top 10 Late-Lactation Phyla Across All Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))

stacked_plot_late


```
See how in Skimmed Milk powder theere is a group of samples where the top ten phyla do not cover close to 100% of the abundance. I will now plot late-lactation phyla without filtering top ten to see which phyla (if there is only one) will fill that gap!

```{r phylum-stacked-bar-ll-full, fig.width=9, fig.height=3.5}


stacked_plot_late_full <- phylum_df_season %>%
  filter(season == "late-lactation" & !is.na(Abundance)) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Late-Lactation Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))
stacked_plot_late_full

```

WOAH! I plotted this and it helped me discover in my two previous plots I was plotting ALL of the samples, just the first plot I was jst showing the ten most abundant phyla in mid-lactation across all samples and then the ten most abundant late-lactation phyla in all samples! Wow, I don't know if I would have caught this otherwise! 

I am now going to make two plots with only the mid-lactation samples (most abundant mid-lactation phyla) and only the late-lactation samples (most abundant late-lactation phyla).


```{r top10-mlphyla-onlyml, fig.width=9, fig.height=3.5}

stacked_plot_mid_onlyml <- phylum_df_season %>%
  filter(season == "mid-lactation" & Phylum %in% top10_phyla_mid) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Top 10 Mid-Lactation Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))

stacked_plot_mid_onlyml


```

Now that looks a little better!

```{r top10-llphyla-onlyll, fig.width=9, fig.height=3.5}

stacked_plot_late_onlyll <- phylum_df_season %>%
  filter(season == "late-lactation" & Phylum %in% top10_phyla_late) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Top 10 Late-Lactation Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))

stacked_plot_late_onlyll

```

Now plot them together

```{r plot-combined-stackedbar, fig.width=9, fig.height=7}

stacked_plot_mid_onlyml + 
stacked_plot_late_onlyll + 
  plot_layout(ncol = 1, guides = "collect") +
  plot_annotation(tag_levels = "A")

```

## Faceted Bar plot