---
title: "Microbial Compositional Analysis of Mid- and Late-lactation Dairy Samples
  Along a Processing Pipeline"
author: "Kailyn Hanke"
date: "2025-04-27"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      fig.path = "../figures/10_Seasonal_Composition/")
```

# Goals 

The goal of this document is to run an analysis of micorbial composition of mid- and late-lactation dairy samples across a processing pipeline, similar to the analysis that I ran in 07_Composition. This is the third of three new analysis documents I will make (I have previously made 08_Seasonal_Biodiversity and 09_Ordination). See more info on my organizational layout under "Goals" in 08_Seasonal_Biodiversity.

1. Load in scaled/normalized phyloseq object. 
2. Calculate the relative abundances of taxonomic groups at various levels: 
    A. Phylum
    B. Genus
    C. ASV 
3. Plot them, and narrow in on specific taxonomic group of interest


## Inputs

1. We will need the `scaled_physeq_season.RData`, which includes a rooted tree that we created in `analysis/09_Seasonal_Ordination/scaled_physeq_season.RData`.

## Outputs 

1. Beautiful visualizations of microbial taxa and how they vary across dairy sample types (including mid- and late-lactation information.
2. Run some stats! 

# Setup 

# Setup 

## Timing of this script

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed

```{r set-seed}
set.seed(111111)
```

## Load Packages 
```{r load-packages}
# Load in the packages 
pacman::p_load(tidyverse, devtools, DT, phyloseq, patchwork, rstatix, 
               install = FALSE)

# load colors
source("code/colors.R")
```

## 1. Load in Scaled Phyloseq object

The following phyloseq object contains microbial community composition data in a standardized format. In this case, we’ve already normalized the reads (scaled to 20090 per sample and including season data), which is essential for comparing relative abundances.

```{r load-data}

load("data/09_Seasonal_Ordination/scaled_physeq_season.RData")

load("data/06_Ordination/scaled_physeq.RData")

load("data/09_Seasonal_Ordination/metadata_df_season.RData")

#Intuition check
min(sample_sums(scaled_physeq)) #19933

#Note for some reason the scaled_phyloseq_season wasn't loading in! Will create scaled_physeq_season again until I fix the error

```

First I will mutate my metadata to include a new column season to define mid- and late-lactation.


```{r update-physeq-df}

#To include seasons, need to update the sample_data in my phyloseq object

# Update the sample_data to include season
sample_data(scaled_physeq) <- sample_data(metadata_df_season)

# Save the updated phyloseq object
scaled_physeq_season <- scaled_physeq

```


# Taxonomic Analysis! 

## A. Phylum 

```{r calc-phylum-df}

phylum_df_season <- 
  scaled_physeq_season %>%
  # Agglomerate all ASV counts within a phylum
  tax_glom(taxrank = "Phylum") %>%
  # Calculate the relative abundance! 
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  # Create a dataframe from phyloseq object
  psmelt() %>%
  #Filter out Phyla < 1 % 
  #dplyr::filter(Abundance > 0.01) %>%
  #Fix the station order 
  mutate(environment_material = fct_relevel(environment_material, c(
    "Bulk Tank Milk", "Collection Tanker Milk", "Whole Milk Silo", "Cream Silo", "Skimmed Milk Silo", "Skimmed Milk Powder")))

#What are the phylum abundances?
phylum_df_season %>%
  group_by(Phylum, season) %>%
  summarize(mean_PercAbund = round(mean(Abundance), digits = 4), .groups = "drop") %>%
  arrange(season, -mean_PercAbund) %>%
  datatable()


top10_phyla_season <- 
  phylum_df_season %>%
  group_by(season, Phylum) %>%
  # group by BOTH season and phylum
  summarize(mean_PercAbund = mean(Abundance), .groups = "drop") %>%
  # regroup by season
  group_by(season) %>%               
  # take top 10 per season
  slice_max(order_by = mean_PercAbund, n = 10) %>%  
  arrange(season, -mean_PercAbund)  

#Identify top phyla in mid-lactation samples
top10_phyla_mid <- 
  phylum_df_season %>%
  filter(season == "mid-lactation") %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = mean(Abundance)) %>%
  arrange(-mean_PercAbund) %>%
  slice_head(n = 10) %>%
  pull(Phylum) #comment out to have mini df w/ abundances

#View
top10_phyla_mid

#Identify top phyla in late-lactation samples
top10_phyla_late <- 
  phylum_df_season %>%
  filter(season == "late-lactation") %>%
  group_by(Phylum) %>%
  summarize(mean_PercAbund = mean(Abundance)) %>%
  arrange(-mean_PercAbund) %>%
  slice_head(n = 10) %>%
  pull(Phylum) #comment out to have mini df w/ abundances


#View
top10_phyla_late

```

## Stacked Bar plots 

### Mid-lactation
```{r phylum-stacked-bar-ml, fig.width=9, fig.height=3.5}

#Intuition check all phyla in top ten for mid-lactation in phylum_colors (in my previous composition doc they were not and it caused me a bunch of avoidable issues)

top10_phyla_mid %in% names(phylum_colors)


#actual plot (by sample)
stacked_plot_mid <- phylum_df_season %>%
  filter(Phylum %in% top10_phyla_mid) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Top 10 Mid-Lactation Phyla Across All Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))

stacked_plot_mid

```


### Late-lactation
```{r phylum-stacked-bar-ll, fig.width=9, fig.height=3.5}

#Intuition check all phyla in top ten for late-lactation in phylum_colors (in my previous composition doc they were not and it caused me a bunch of avoidable issues)

top10_phyla_late %in% names(phylum_colors)



#actual plot (by sample)
stacked_plot_late <- phylum_df_season %>%
  filter(Phylum %in% top10_phyla_late) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Top 10 Late-Lactation Phyla Across All Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))

stacked_plot_late


```
See how in Skimmed Milk powder there is a group of samples where the top ten phyla do not cover close to 100% of the abundance. I will now plot late-lactation phyla without filtering top ten to see which phyla (if there is only one) will fill that gap!

```{r phylum-stacked-bar-ll-full, fig.width=9, fig.height=3.5}


stacked_plot_late_full <- phylum_df_season %>%
  filter(season == "late-lactation" & !is.na(Abundance)) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Late-Lactation Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))
stacked_plot_late_full

```

WOAH! I plotted this and it helped me discover in my two previous plots I was plotting ALL of the samples, just the first plot I was jst showing the ten most abundant phyla in mid-lactation across all samples and then the ten most abundant late-lactation phyla in all samples! Wow, I don't know if I would have caught this otherwise! 

I am now going to make two plots with only the mid-lactation samples (most abundant mid-lactation phyla) and only the late-lactation samples (most abundant late-lactation phyla).


```{r top10-mlphyla-onlyml, fig.width=9, fig.height=3.5}

stacked_plot_mid_onlyml <- phylum_df_season %>%
  filter(season == "mid-lactation" & Phylum %in% top10_phyla_mid) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Top 10 Mid-Lactation Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))

stacked_plot_mid_onlyml


```

Now that looks a little better!

```{r top10-llphyla-onlyll, fig.width=9, fig.height=3.5}

stacked_plot_late_onlyll <- phylum_df_season %>%
  filter(season == "late-lactation" & Phylum %in% top10_phyla_late) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(limits = c(0, 1)) +
  facet_grid(. ~ environment_material, scales = "free_x", space = "free_x") +
  labs(title = "Top 10 Late-Lactation Phyla",
       y = "Relative Abundance", x = "Sample ID") +
  scale_fill_manual(values = phylum_colors) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        panel.spacing.x = unit(0.5, "lines"))

stacked_plot_late_onlyll

```

Now plot them together

```{r plot-combined-stackedbar, fig.width=9, fig.height=7}

stacked_plot_mid_onlyml + 
stacked_plot_late_onlyll + 
  plot_layout(ncol = 1, guides = "collect") +
  plot_annotation(tag_levels = "A")

```

top10_phyla_mid

 [1] "Bacillota"               "Pseudomonadota"          "Bacteroidota"           
 [4] "Actinomycetota"          "Deinococcota"            "Verrucomicrobiota"      
 [7] "Spirochaetota"           "Thermodesulfobacteriota" "Acidobacteriota"        
[10] "Planctomycetota"         "Myxococcota"

Plus "Myxococcota" (late) and Deinococcota is (mid)

## Faceted Bar plot

Now I would like to make a faceted bar plot with the top 10 phyla of both mid-lactation and late-lactation samples. In the top ten phyla for each season, 9 are the same. In the mid-lactation samples Deinococcota makes the top ten, while in the late-lactation samples, Myxococcota makes the cut. 

So in this graph I will plot the 9 shared top 10 phyla, and then both Deinococcota and Myxococcota (total of 11 phyla). I will also have a bar for each season within each phyla plot.

```{r phylum-facet-bar, fig.width=5, fig.height=12}


phylum_df_season %>%
  filter(Phylum %in% c(
    "Bacillota", "Pseudomonadota", "Bacteroidota", "Actinomycetota", "Verrucomicrobiota", "Spirochaetota", "Thermodesulfobacteriota",
    "Acidobacteriota", "Planctomycetota", "Deinococcota", "Myxococcota"
  )) %>%
  ggplot(aes(x = season, y = Abundance, fill = Phylum, color = Phylum)) + 
  facet_wrap(~Phylum, scales = "free", nrow = 3) + 
  geom_jitter(width = 0.2, alpha = 0.5) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.5) + 
  scale_fill_manual(values = phylum_colors) + 
  scale_color_manual(values = phylum_colors) + 
  theme_bw() + 
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
  ) +
  labs(x = "Season", y = "Relative Abundance", title = "Phylum Abundance by Season")

```

```{r add-stats-11phyla}


# Do Wilcoxon test for each Phylum
wilcox_results <- phylum_df_season %>%
  filter(Phylum %in% c(
    "Bacillota", "Pseudomonadota", "Bacteroidota", "Actinomycetota", "Verrucomicrobiota", "Spirochaetota", "Thermodesulfobacteriota",
    "Acidobacteriota", "Planctomycetota", "Deinococcota", "Myxococcota"
  )) %>%
  group_by(Phylum) %>%
  wilcox_test(Abundance ~ season) %>%
  adjust_pvalue(method = "fdr") %>%  # Optional: adjust for multiple testing (FDR)
  add_significance("p.adj")


#fix sig plot

# Calculate the maximum Abundance for each Phylum
phylum_max <- phylum_df_season %>%
  group_by(Phylum) %>%
  summarise(max_abundance = max(Abundance, na.rm = TRUE)) %>%
  mutate(y.position = 1.05 * max_abundance)  # Add 5% extra space


# Assuming your wilcox_results has a 'Phylum' column
wilcox_results2 <- wilcox_results %>%
  left_join(phylum_max, by = "Phylum")

# Plot
phylum_df_season %>%
  filter(Phylum %in% c(
    "Bacillota", "Pseudomonadota", "Bacteroidota", "Actinomycetota", "Verrucomicrobiota", "Spirochaetota", "Thermodesulfobacteriota",
    "Acidobacteriota", "Planctomycetota", "Deinococcota", "Myxococcota"
  )) %>%
  ggplot(aes(x = season, y = Abundance, fill = Phylum, color = Phylum)) + 
  facet_wrap(~Phylum, scales = "free", nrow = 3) + 
  geom_jitter(width = 0.2, alpha = 0.5) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.5) + 
  scale_fill_manual(values = phylum_colors) + 
  scale_color_manual(values = phylum_colors) + 
  theme_bw() + 
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
  ) +
  labs(x = "Season", y = "Relative Abundance", title = "Phylum Abundance by Season") +
  stat_pvalue_manual(
  wilcox_results2,
  label = "p.adj.signif",
  y.position = "y.position",   # <-- dynamic y-positions!
  bracket.size = 0.5,
  tip.length = 0.01,
  size = 5
)





```


Now we will focus in on certain phyla of interest


### Deinococcota


```{r phylum-deinococcota, fig.width=7, fig.height=3.5}

deinococcota_phylum <- 
  phylum_df_season %>%
  filter(Phylum == "Deinococcota") %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Deinococcota Phylum", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.position = "right"
  )

deinococcota_phylum
```


### Myxococcota

DID NOT RUN STATS AT ALL - CHECK 284


```{r phylum-myxococcota, fig.width=7, fig.height=3.5}

myxococcota_phylum <- 
  phylum_df_season %>%
  filter(Phylum == "Myxococcota") %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Myxococcota Phylum", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.position = "right"
  )

myxococcota_phylum

```

### Bacillota


```{r phylum-bacillota, fig.width=7, fig.height=3.5}

bacillota_phylum <- 
  phylum_df_season %>%
  filter(Phylum == "Bacillota") %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Bacillota Phylum", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.position = "right"
  )

bacillota_phylum

```

### Bacteroidota


```{r phylum-bacteroidota, fig.width=7, fig.height=3.5}

bacteroidota_phylum <- 
  phylum_df_season %>%
  filter(Phylum == "Bacteroidota") %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Bacteroidota Phylum", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.position = "right"
  )

bacteroidota_phylum

```

### Pseudomonadota


```{r phylum-pseudomonadota, fig.width=7, fig.height=3.5}

pseudomonadota_phylum <- 
  phylum_df_season %>%
  filter(Phylum == "Pseudomonadota") %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Pseudomonadota Phylum", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.position = "right"
  )

pseudomonadota_phylum

```

## Genus

Now time to follow up with some deeper taxonomic analysis at the genus level.

-Deinococcota
-Bacteroidota

```{r genus-df}

genus_df_season <- 
  scaled_physeq_season %>%
  # Agglomerate all ASV counts within a phylum
  tax_glom(taxrank = "Genus") %>%
  # Calculate the relative abundance! 
  transform_sample_counts(function(x) {x/sum(x)}) %>%
  # Create a dataframe from phyloseq object
  psmelt() %>%
  #Filter out Phyla < 1 % 
  #dplyr::filter(Abundance > 0.01) %>%
  #Fix the station order 
  mutate(environment_material = fct_relevel(environment_material, c(
    "Bulk Tank Milk", "Collection Tanker Milk", "Whole Milk Silo", "Cream Silo", "Skimmed Milk Silo", "Skimmed Milk Powder")))

```

### Deinococcota Genera

```{r deino-genus, fig.width=6, fig.height=6}

deino_genus <- 
  genus_df_season %>%
  filter(Phylum == "Deinococcota") %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  facet_wrap(~Genus, scales = "free_y", nrow = 1) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Deinococcota Genera", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    legend.position = "right"
  )

# View it
deino_genus


```

### Bacteroidota Genera

```{r bacteroid-genus, fig.width=9, fig.height=6}

#See how many genera in Bacteroidota because I am curious
genus_df_season %>% 
  filter(Phylum == "Bacteroidota") %>% 
  distinct(Genus) %>% 
  nrow()

#wow there are 195 genera, we simply cannot plot them all!

#instead I will plot the 10 Bacteroidota genera with the highest mena abundance across all samples/seasons

# Find top 10 Bacteroidota genera (highest mean abundance across all seasons/samples)
top10_bacteroidota_genera <- genus_df_season %>%
  filter(Phylum == "Bacteroidota") %>%
  group_by(Genus) %>%
  summarize(mean_abund = mean(Abundance, na.rm = TRUE)) %>%
  arrange(desc(mean_abund)) %>%
  slice_head(n = 10) %>%
  pull(Genus)

# View it!
top10_bacteroidota_genera


# Now plot only those top 10
bacteroid_genus <- 
  genus_df_season %>%
  filter(Phylum == "Bacteroidota", Genus %in% top10_bacteroidota_genera) %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  facet_wrap(~Genus, scales = "free_y", nrow = 2) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Top 10 Bacteroidota Genera", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# View it
bacteroid_genus

```

# ASV Level

```{r ASV-df}

# Note: The read depth MUST be normalized in some way: scale_reads 
ASV_df_season <- 
  scaled_physeq_season %>%
  # Prune out ASVs that have fewer than 100 counts! 
  ## LOOK AT HOW MANY ARE REMOVED! We scaled to 20090 reads! 
  prune_taxa(taxa_sums(.) >= 2009, .) %>%
  # agglomerate at the phylum level 
  tax_glom(taxrank = "ASV") %>% 
  # Transform counts to relative abundance 
  transform_sample_counts(function (x) {x/sum(x)}) %>%
  # Melt to a long format 
  psmelt() %>%
  # fix the order of date
  mutate(environment_material = fct_relevel(environment_material, c(
    "Bulk Tank Milk", "Collection Tanker Milk", "Whole Milk Silo", "Cream Silo", "Skimmed Milk Silo", "Skimmed Milk Powder")))

#View
ASV_df_season

```

### Deinococcota ASVs

```{r deino-asvs, width=6, fig.height=8}

#Calculate top ASVs (mean abundance > 0.005)

top_deino_ASVs <- 
  ASV_df_season %>%  
  filter(Phylum == "Deinococcota") %>%
  group_by(ASV) %>%
  summarize(mean_Abundance = mean(Abundance, na.rm = TRUE)) %>%
  filter(mean_Abundance > 0.005) %>%
  pull(ASV)

#Plot
deino_asv_sample_type <- 
  ASV_df_season %>%  
  filter(ASV %in% top_deino_ASVs) %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  facet_wrap(Genus~ASV, scales = "free_y", nrow = 2) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Top Deinococcota ASVs", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# View it
deino_asv_sample_type


```

### Bacteroidota ASVs

```{r bacteroid-asvs, fig.width=6, fig.height=8}

#Calculate top ASVs (mean abundance > 0.005)

top_bacteroid_ASVs <- 
  ASV_df_season %>%  
  filter(Phylum == "Bacteroidota") %>%
  group_by(ASV) %>%
  summarize(mean_Abundance = mean(Abundance, na.rm = TRUE)) %>%
  filter(mean_Abundance > 0.005) %>%
  pull(ASV)

#Plot
bacteroid_asv_sample_type <- 
  ASV_df_season %>%  
  filter(ASV %in% top_bacteroid_ASVs) %>%
  ggplot(aes(x = environment_material, y = Abundance, 
             fill = season, color = season)) + 
  facet_wrap(Genus~ASV, scales = "free_y", nrow = 2) + 
  geom_boxplot(position = position_dodge(width = 0.7), alpha = 0.5, outlier.shape = NA) + 
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.7), alpha = 0.7) + 
  theme_bw() + 
  labs(title = "Top Bacteroidota ASVs", x = "Sample Type", y = "Relative Abundance") + 
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# View it
bacteroid_asv_sample_type

```


### Identifying the 10 most abundant ASVs:

```{r plot-top-asvs}

#identify my top ten ASVs total (all samples)
top_asvs_total <- names(sort(taxa_sums(scaled_physeq_season), decreasing = TRUE))[1:10]

#View my top 10 ASVs
top_asvs_total


# Mid-lactation
top_asvs_mid <- 
  scaled_physeq_season %>%
  # Filter for mid-lactation samples
  subset_samples(season == "mid-lactation") %>% 
   # Get the abundance for each ASV
  taxa_sums() %>% 
  sort(decreasing = TRUE) %>%
  names() %>%
  # Get the top 10
  .[1:10]  

# View the top 10 ASVs for mid-lactation
top_asvs_mid



# Late-lactation
top_asvs_late <- 
  scaled_physeq_season %>%
  # Filter for late-lactation samples
  subset_samples(season == "late-lactation") %>% 
   # Get the abundance for each ASV
  taxa_sums() %>% 
  sort(decreasing = TRUE) %>%
  names() %>%
  # Get the top 10
  .[1:10]  

# View the top 10 ASVs for mid-lactation
top_asvs_late


```

### Plotting most abundant ASVs



```{r plot-most-abundant}

#Plot top 10 abundant ASVs across all samples

ASV_df_season %>%
  filter(ASV %in% top_asvs_total) %>%
  ggplot(aes(x = environment_material, y = Abundance,
             fill = season, color = season)) +  # Map 'season' to fill and color
  facet_wrap(Phylum ~ ASV, scales = "free_y", nrow = 2) +  # Facet by Phylum and ASV
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +  # Add boxplot
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), 
             alpha = 0.7, size = 2) +  # Adjust jitter for separation
  theme_bw() +
  labs(title = "Top 10 Total ASVs by Season") +
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) +
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))


#Plot top ten most abundant ASVs in mid-lactation samples

ASV_df_season %>%
  filter(ASV %in% top_asvs_mid) %>%
  ggplot(aes(x = environment_material, y = Abundance,
             fill = season, color = season)) +  # Map 'season' to fill and color
  facet_wrap(Phylum ~ ASV, scales = "free_y", nrow = 2) +  # Facet by Phylum and ASV
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +  # Add boxplot
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), 
             alpha = 0.7, size = 2) +  # Adjust jitter for separation
  theme_bw() +
  labs(title = "Top 10 Mid-Lactation ASVs by Season") +
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) +
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))


#Plot top ten most abundant ASVs in late-lactation samples

ASV_df_season %>%
  filter(ASV %in% top_asvs_late) %>%
  ggplot(aes(x = environment_material, y = Abundance,
             fill = season, color = season)) +  # Map 'season' to fill and color
  facet_wrap(Phylum ~ ASV, scales = "free_y", nrow = 2) +  # Facet by Phylum and ASV
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +  # Add boxplot
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), 
             alpha = 0.7, size = 2) +  # Adjust jitter for separation
  theme_bw() +
  labs(title = "Top 10 Late-Lactation ASVs by Season") +
  scale_color_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) +
  scale_fill_manual(values = c("mid-lactation" = "skyblue", "late-lactation" = "tomato")) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))

```

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```
